Embedding to Reference t-SNE Space Addresses Batch Effects in Single-Cell Classification

### 2019-08-13 paper/policar2019-tsne-embedding.tex
Minor edits.

\begin{figure}[htb]
\includegraphics[width=\textwidth]{figures/transform_brain.pdf}
\caption{A two-dimensional embedding of a reference containing brain cells (a) and
the corresponding mapping of secondary data containing hypothalamic cells (b). Notice
that the majority of hypothalamic cells were mapped to their corresponding
reference cluster. For instance, astrocyte cells marked with red on the
right were mapped to an oval cluster of same-typed cells denoted with the same
color in the visualization on the left.} \label{fig:transform_brain}
\end{figure}

@@

We apply the proposed approach to t-SNE visualizations of
single-cell data. Data in this realm include a variety of
cells from specific tissues, and characterize them through gene expression.
In our experiments, we considered several recently published data
sets where cells were annotated with the cell type. Our aim was to construct
t-SNE visualizations where similarly-typed cells would cluster together,
despite systematic differences between data sources. Below, we list the data sets,
describe single-cell specific data preprocessing procedures, and display the
resulting data visualizations. Finally, we discuss the success of the proposed
approach in alleviating the batch effects.

@@

In principle, our theoretically-grounded embedding of secondary data into the
scaffold defined by the reference embedding could be simplified with the
application of the nearest neighbors-based procedure. For example, while
describing a set of tricks for t-SNE, Kobak and Berens~\cite{art_of_using_tsne}
proposed positioning new points into a known embedding by placing them in the
median position of their 10 nearest neighbors, where the neighborhood was
estimated in the original data space. Notice that we use this approach as well,
but only for the initialization of positions of new data instances that are subject
to further optimization. In Fig.~\ref{fig:optimization} we demonstrate that
nearest neighbor-based positioning is insufficient and may yield clumped
visualizations where the optimal positioning using the t-SNE loss function is
much more dispersed and rightfully shows a more considerable variation in the
secondary data. Some data points may also fall into the empty regions
between different clusters, while after optimization they typically
move closer to same-typed groups. While the nearest neighbour-based approach
always places new data points to their most similar cluster, our proposed
t-SNE approach can identify extreme outliers and place them far from the
reference cluster.

@@

Our procedure and the resulting cell landscape are affectected by the asymmetry
of determining the reference and the secondary datasets. In practice, however,
newly produced \em secondary \em datasets would be embedded into published and
annnotated the reference landscapes. Scaling-up to a wider collection of cell
types is driven by increasingly large collections of data, e.g the Human Cell
Atlas initiative~\cite{hca}.

>>>>

\begin{figure}[htb]
\includegraphics[width=\textwidth]{figures/transform_brain.pdf}
\caption{A two-dimensional embedding of a reference containing brain cells (a) and
the corresponding mapping of secondary data containing hypothalamic cells (b). 
The majority of hypothalamic cells were mapped to their corresponding
reference cluster. For instance, astrocyte cells marked with red on the
right were mapped to an oval cluster of same-typed cells denoted with the same
color in the visualization on the left.} \label{fig:transform_brain}
\end{figure}

@@

We apply the proposed approach to t-SNE visualizations of
single-cell data. Data in this realm include a variety of
cells from specific tissues, and characterize them through gene expression.
In our experiments, we considered several recently published data
sets where cells were annotated with the cell type. Our aim was to construct
t-SNE visualizations where similarly-typed cells would cluster together,
despite systematic differences between data sources. To that end, we focus
on comparing different ways of using t-SNE rather than differences to embeddings like PCA or MDS, which
have been substantially covered before~\todo{Add citation}. 
Below, we list the data sets,
describe single-cell specific data preprocessing procedures, and display the
resulting data visualizations. Finally, we discuss the success of the proposed
approach in alleviating the batch effects.

@@

In principle, our theoretically-grounded embedding of secondary data into the
scaffold defined by the reference embedding could be simplified with the
application of the nearest neighbors-based procedure. For example, while
describing a set of tricks for t-SNE, Kobak and Berens~\cite{art_of_using_tsne}
proposed positioning new points into a known embedding by placing them in the
median position of their 10 nearest neighbors, where the neighborhood was
estimated in the original data space. Notice that we use this approach as well,
but only for the initialization of positions of new data instances that are
subject to further optimization. Despite both nearest-neighbours search and
t-SNE optimization can be computed in linear time, the former dominates the
runtime (mouse retina example; M reference, N secondary cells, X min N-search,
Y min optimization).

We demonstrate a case where nearest neighbor-based positioning is insufficient
in Fig.~\ref{fig:optimization}. It may yield clumped visualizations where the
optimal positioning using the t-SNE loss function is much more dispersed and
rightfully shows a more considerable variation in the secondary data. 
Some data points may also fall into the empty regions
between different clusters, while after optimization they typically
move closer to same-typed groups. While the nearest neighbour-based approach
always places new data points to their most similar cluster, our proposed
t-SNE approach can identify extreme outliers and place them far from the
reference cluster.

@@

Our procedure and the resulting cell landscape are affectected by the asymmetry
of determining the reference and the secondary datasets. In practice, however,
newly produced \em secondary \em data would be embedded into published and
annotated reference landscapes. Scaling-up to a wider collection of cell
types is driven by increasingly large collections of data, e.g the Human Cell
Atlas initiative~\cite{hca}.
